Transactions Fraud Detection using Machine Learning and Nature Inspired Algorithms

### 2020-05-29 paper/main.tex
Write experiments and conclusion sections

@@

@@

>>>>

\usepackage{booktabs}
\usepackage{subfig}

@@

We did several experiemnts in which we compared the following algorithms:

\begin{itemize}
	\item Firefly Algorithm \cite{fister2013comprehensive} - FA (alpha = 1, betamin = 1, gamma = 2)
	\item Cuckoo Search \cite{yang2009cuckoo} - CS (pa = 0.2, alpha = 0.5)
	\item Bat Algorithm \cite{yang2010new} - BA (A = 0.5, r = 0.5, Qmin = 0.0, Qmax = 2.0)
	\item Flower Pollination Algorithm \cite{yang2012flower} - FPA (p = 0.35, beta = 1.5)
	\item Grey Wolf Optimizer \cite{Mirjalili_Mirjalili_Lewis_2014} - GWO
\end{itemize}

As a baseline models we have chosen the model with all features and model with selected features from \textit{recursive feature selection with cross-validation}.

\subsection{Experiments strategy}

We chose a \textit{decision tree classifier} as a model on which we will try to find the optimal features. We kept the default hyper-parameters for this classifier set by scikit-learn library.

Since the dataset, if pretty large, firstly we \textit{undersampled} it to achieve the equal size of both classes. Then we shrunk it again to 5\% of its size, so we would be able to make as many experiments as possible (we did not have enough computing power). We did a train/test split of 80/20.

For each algorithm, we tried several population sizes (10, 20, and 50), for which we did 5 runs of the algorithm. We trained exactly 50 generations. For both, recursive feature selection and feature selection using nature-inspired algorithms we did 5 fold cross-validation. We did not experiment with algorithm parameters and we used their defaults which are described in the previous section.

\subsection{Results}

When comparing the population sizes, all of the algorithms achieved the best results when the population size was highest (50). The only exception is Cuckoo Search, which found the most optimal solution with population size 10, however the difference negligible. Table \ref{tab:nia_comparison}. shows the algorithm scores for each population. Figure \ref{fig:nia_train_score_by_algorithm}. shows the differences in scores between population size of 10 and 50.

\begin{figure}
    \centering
    \subfloat{{ 
        \includegraphics[width=5.5cm]{figures/nia_train_score_by_algorithm_10}
    }}
    \qquad
    \subfloat{{
        \includegraphics[width=5.5cm]{figures/nia_train_score_by_algorithm_50}
    }}
    \caption{Scores of the algorithms by the population size. The scores are slightly higher for population size of 50.}
    \label{fig:nia_train_score_by_algorithm}
\end{figure}

\begin{table}[]
	\begin{center}
		\begin{tabular}{|l|rrrrr|}
			\hline
			\textbf{Population size \textbackslash Score} & \multicolumn{1}{l}{\textbf{FA}} & \multicolumn{1}{l}{\textbf{CS}} & \multicolumn{1}{l}{\textbf{BA}} & \multicolumn{1}{l}{\textbf{FPA}} & \multicolumn{1}{l|}{\textbf{GWO}} \\ \hline
			\textbf{10}                                             & 0.753                           & \textbf{0.752}                           & 0.759                           & 0.754                            & 0.776                             \\
			\textbf{20}                                             & 0.757                           & 0.746                           & 0.755                           & 0.750                            & 0.771                             \\
			\textbf{50}                                             & \textbf{0.761}                           & 0.751                           & \textbf{0.761}                           & \textbf{0.756}                            & \textbf{0.782}                             \\ \hline
		\end{tabular}
	\end{center}
	\begin{center}
		\caption{Highest train scores (ROC AUC) of nature inspired algorithms by population size. We may notice that Grey Wolf Algorithm (GWO) outperformed all of the other algorithms.} \label{tab:nia_comparison}
	\end{center}
\end{table}

All of the nature-inspired algorithms perform roughly the same, except the Grey Wolf Optimizer which outperformed the others (Figure \ref{fig:nia_train_score_by_algorithm}). These differences in results (means) between GWO and other algorithms are significant (t-test; $p-value=0.012$). Looking for the explanation, we found out that only Grey Wolf Algorithm and Bat Algorithm converged during the optimization run. Figure \ref{fig:nia_scores_by_run} shows the optimization of the Firefly Algorithm and Grey Wolf Optimizer, the difference is noticeable. If we fine-tuned the parameters of other algorithms, they might converge as well. The fact that GWO does not have any parameters (except the population size) was an advantage for it in our experiments.

% https://towardsdatascience.com/hypothesis-testing-in-machine-learning-using-python-a0dc89e169ce

\begin{figure}
    \centering
    \subfloat[Firefly Algorithm]{{ 
        \includegraphics[width=5.5cm]{figures/fa_scores_by_run}
    }}
    \qquad
    \subfloat[Grey Wolf Optimizer]{{
        \includegraphics[width=5.5cm]{figures/gwo_scores_by_run}
    }}
    \caption{Optimization progress for 5 runs of Grey Wolf Optimizer and Firefly Algorithm. The Firefly Algorithm did not really converge.}
    \label{fig:nia_scores_by_run}
\end{figure}

The best features, selected by the optimization process, were later used to train the final model on test data. These models were compared to a model trained on all features and the model trained on features selected by recursive feature elimination.

The nature-inspired algorithms found a more optimal set of features on the train dataset than recursive feature elimination. However, they achieved worse results on the test dataset. The only exception is Grey Wolf Optimizer, which achieved the best train and test score. The most noticeable difference in the results is the number of selected features by algorithms. All of the algorithms (including recursive feature elimination), selected around $~110$ features, however, GWO selected only 53, this is probably the reason of its results. These results are shown in Table \ref{tab:nia_baseline_comparison}. Figure \ref{fig:nia_selected_features_by_algorithm} shows that GWO selected fewer features in general not only in its best solution.

\begin{table}[]
	\begin{center}
		\begin{tabular}{|lrrr|}
			\hline
			\textbf{Feature selection type} & \textbf{Train score} & \textbf{Test score} & \textbf{Selected features} \\ \hline
			All features                    & N/A                  & 0.713               & 240                                  \\
			Recursive feature elimination   & 0.747                & 0.719               & 112                                  \\
			Firefly Algorithm               & 0.761                & 0.663               & 114                                  \\
			Cuckoo Search                   & 0.753                & 0.718               & 106                                  \\
			Bat Algorithm                   & 0.761                & 0.689               & 125                                  \\
			Flower Pollination Algorithm    & 0.756                & 0.706               & 113                                  \\
			\textbf{Grey Wolf Optimizer}             & \textbf{0.782}                & \textbf{0.733}               & \textbf{53}                                   \\ \hline
		\end{tabular}
	\end{center}
	\begin{center}
		\caption{Comparison of the baseline models with nature-inspired algorithms. The train score stands for the best score from the optimization. All of the nature-inspired algorithms (except Grey Wolf Optimizer) achieved a better train score than recursive feature elimination, however, they overfit the model which resulted in a worse test scores.} \label{tab:nia_baseline_comparison}
	\end{center}
\end{table}

\begin{figure}[ht]
	\begin{center}
	    \includegraphics[width=5.5cm]{figures/nia_selected_features_by_algorithm.png}
    \end{center}
	\caption{Number of selected features by algorithm - Grey Wolf Algorithm selected by far the least number of features.}
	\label{fig:nia_selected_features_by_algorithm}
\end{figure}

@@

In our experiments, we found out Grey Wolf Optimizer as the best nature-inspired algorithm for feature selection. It achieved significantly better results than other nature-inspired algorithms and better results that recursive feature elimination.

GWO alongside Bat Algorithm was the only nature-inspired algorithms that were able to converge during the optimization process. We assume that other algorithms might converge too if we experimented with their parameter settings. We consider the fact, that GWO does not have any parameters other than population size as a big strength, because it saves a significant amount of optimization time.

In comparison with traditional ways of feature selection (recursive feature elimination), all of the nature-inspired algorithms were able to find better solutions on the training dataset, however, they overfit most of the time. Moreover, the feature selection using nature-inspired algorithms was much slower.
